\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-15 response.tex

\begin{itemize}
\item graphs drawn on a fixed surface with each edge in $\leq k$ crossings,
\item map graphs of degree $\leq k$ graphs embedded in a fixed surface,
\item string graphs with each curve in $\leq k$ intersections,
\item $k$-nearest neighbour graphs in the plane, and
\item powers of classes of graphs of degree $\leq k$ which admit a product structure theorem.
\end{itemize}

All but the last bullet point are shown to follow from the first.
The main results in the paper are correct, strong, and of high interest,
and the paper should definitely be accepted to JCTB after some
(fairly significant) revisions. In particular, the proof of Theorem 3 is
difficult to read due to the notation, and the paper could be substantially
streamlined without losing the above results. Detailed suggestions follow.

Major comments:

The following comments are all related to the proof of Theorem~3.  Much of the difficulty comes from the fact that there are many things to keep track of. We are a given a graph $G$, a shortcut system $\mathcal{S}$, a layering $\mathcal{L}$, a vertex-partition $\mathcal{P}$, and a tree decomposition $\mathcal{T}$ of the quotient graph $H:=G/\mathcal{P}$.  Now we have to use these to show the existence of a new layering $\mathcal{L}'$, a new vertex partition $\mathcal{P'}$, a new tree decomposition $\mathcal{T}'$ of the new quotient graph $H':=G^{\mathcal{S}}$.

During the proof, a number of related sets come up, which we have given names to. As a referee point out, these could remain unnamed and be described using an appropriate notation.

\todo[inline]{Write something more here depending on how the experiment with new notation works out.}

A secondary issue is that a number of the proofs in this section are essentially "definition checking".  The relationships between the various vertex sets that arise in the proof really do just follow from their definitions (which were made with this purpose in mind).  We have done this checking explicitly, though each of these proofs could also be replaced by a sentence of the form ``This [claim/lemma] follows from the definitions of\ldots''.  It is not clear what the best choice is.

The real action begins in the proof Theorem 9 (using numbering from the original submission) and especially Claim 3 (which involves some careful counting) and Claim 5 (which involves checking that we have everything needed to apply the Pillipczuk-Siebertz lemma on $k$-reachability in ``upward-directed'' graphs of treewidth $t$.)

\todo[inline]{Write more here depending on how the notation experiment turns out.}

 % and new vertex partition of a related graph $G^\mathcal{S}$ as we  A large part of the

### 2022-03-15 k-planar.tex
Pfffft

@@

  To establish Property~\yref{ancestor-edge}, consider some vertex $w\in F_x$.  Since $w\in F_x$, there exists an edge $vw\in E(G)$ with $v\in V_x$ and $w\not\in V_x$.  Since $v\in V_x$, $v\in Y_{x'}$ for some $T$-descendant $x'$ of $x$ (possibly $x=x'$). Since $\mathcal{Y}$ is a partition, $w\in Y_{a}$ for some $a\not\in V(T_x)$.  Since $vw\in E(G)$, we have $x'a\in E(H)$.  By \tref{ancestor-edge}, one of $a$ or $x'$ is a $T$-ancestor of the other. Since $w\in Y_a\subseteq V_a$ and $w\not\in V_x\supseteq V_{x'}$, \yref{containment-i} rules out the possibility that $x'$ is a $T$-ancestor of $a$. Therefore, $a$ is a $T$-ancestor of $x$ which is a $T$-ancestor of $x'$.  Let $z_0,\ldots,z_r$ be the path in $T$ from $z_0:=x'$ to $z_r:=a$.  For each $i\in\{0,\ldots,r\}$, at least one of $a$ or $x'$ is in $B_{z_i}$, since $x'a\in E(H)$.  However, by \tref{subtree-root} $x'$ is not contained in $B_{x_i}$ for any $i\in\{1,\ldots,r\}$.  Therefore $a\in B_{x_i}$ for each $i\in\{0,\ldots,r\}$.  In particular, $a$ is contained in $B_x$.
  Property~\yref{ancestor-edge} now follows from the fact that $|B_x|\le t+1$ and $B_x$ contains $x$.
\end{proof}

\referee{2}{Page 7.  In the very last sentence, I think $B_{x_i}$ should be $B_{z_i}$
(twice).  Moreover, the way the proof is written, it seems to suggest
that we only get that $a \in B_{z_i}$ for each $i \in \{1, \dots, r\}$.
The conclusion that $a \in B_{z_i}$ for each $i \in \{0, \dots, r\}$ is
true though.  In a normalized tree-decomposition, if $xy$ is an edge and
$x$ is a $T$-ancestor of $y$, then $x$ must be in $B_y$ (since the subtrees $T_x$
and $T_y$ intersect).}

@@

@@

@@

\referee{2}{Page 18.  Say $H'$ is connected since $X$ is connected.}

@@

\note{DW}{Say something about how \cref{1_planar_is_4_framed} is better than similar results in the literature.}

@@

\begin{thm}
\label{gkPlanarProduct}
Every $(g,k)$-planar graph is contained in $H\boxtimes P \boxtimes K_\ell$ for some graph $H$ with $\tw(H) \leq \binom{k+4}{3}-1$, where $\ell:=\max\{2g,3\}\cdot(6k^2+16k+10)$.
\note{DW}{I have changed $\tw(H) \leq \binom{k+5}{4}-1$ to $\tw(H) \leq \binom{k+4}{3}-1$, okay?}
\end{thm}

\note{DW}{I am pretty sure one can get treewidth 3 for $(g,1)$-planar graphs, combining \cref{framed} with the method in \citep{DHHW}. I think the best place for this result is reference \citep{DHHW} since the present paper pre-dates \citep{DHHW} (and this addition will increase the significance of \citep{DHHW}). Okay?}

@@

\note{DW}{I don't want to do a complete restructuring. In our response, let's just say why we structure it the way we do.}

@@

\cref{qn,gkPlanarProduct} imply that $(g,k)$-planar graphs have queue-number at most $g 2^{O(k^3)}$. \note{DW}{I have changed $g 2^{O(k^4)}$ to $g 2^{O(k^3)}$, okay?}
Note that \citet{DJMMUW20} previously proved the bound of
$O(g^{k+2})$ using \cref{GenusProduct} and an ad-hoc method. Our result provides a better bound when $g>2^{k^2}$ \note{DW}{I have changed $g>2^{k^3}$ to $g>2^{k^2}$, okay?}. In the case of 1-planar graphs we can improve further. \citet{ABGKP20} proved that every planar graph with treewidth at most $3$ has queue-number at most $5$. Thus the graph $H$ in \cref{1-planar} has queue-number at most $5$. \cref{qn,1-planar} then imply:

@@

\begin{cor}\quad
\begin{compactitem}
\item	For every $1$-planar graph $G$, $\pi(G)\le 7\times 4^4=1792$.
%\item For every $k$-planar graph $G$,
%	$\pi(G)\le (18k^2+48k+30) 4^{\binom{k+4}{3}}$.
\item	For every $(g,k)$-planar graph $G$,
	$  \pi(G)\le \max\{2g,3\}\cdot(6k^2+16k+10) 4^{\binom{k+4}{3}}.$
\note{DW}{I have changed $4^{\binom{k+5}{4}}$ to $4^{\binom{k+4}{3}}$, okay?}
\end{compactitem}
\end{cor}

@@

\begin{cor}\quad
\begin{compactitem}
\item For every $1$-planar graph $G$,\; $ \chi_p(G)\le 5 (p+3)(p+2)(p+1)^2$.
%\item For every $k$-planar graph $G$,\; $\displaystyle \chi_p(G)\le (18k^2+48k+30)(p+1) \binom{p+ \binom{k+4}{3}-1}{ \binom{k+4}{3}-1}$.
\item For every $(g,k)$-planar graph $G$,\;
$\displaystyle \chi_p(G) \le \max\{2g,3\}\cdot(6k^2+16k+10) (p+1) \binom{p+\binom{k+4}{3}-1}{\binom{k+4}{3}-1}$.
\note{DW}{I have changed $\binom{k+5}{4}$ to $\binom{k+4}{3}$, okay?}
 \end{compactitem}
\end{cor}

@@

\note{DW}{This subsection needs a clean-up; define $(g,d)$-framed graph. Is the following lemma known (at least in the $g=0$ case)?}

\begin{lem}
\label{MewMapGraphLemma}
for all integers $g\geq 0$ and $d\geq 3$, every $(g,d)$-map graph is a spanning subgraph of  $(g,d)$-framed graph.
\end{lem}

@@

\note{DW}{\cref{MapPartition} has been updated.}

\note{DW}{I am pretty sure one can get treewidth 3 here using \cref{framed} and the method in \citep{DHHW}. I think the best place for this result is reference \citep{DHHW} since the present paper pre-dates \citep{DHHW} (and this addition will increase the significance of \citep{DHHW}). Okay?}

@@

\note{DW}{\cref{StringPartition} has been updated.}

@@

\note{DW}{\cref{PowerGenus} has been updated}

>>>>

\note{PM}{I've added to the \texttt{response.tex} file explaining some of this.  Each of (Y1), (Y3), (Y4), and (Y5) have a one-sentence justification. The justification for (Y2) is just definition unpacking and then using the observation that the referee recommends.}

\note{PM}{I plan to do an experiment with the referee's suggested notation, in a separate git branch to see how it turns out.}

@@

  To establish Property~\yref{ancestor-edge}, consider some vertex $w\in F_x$.  Since $w\in F_x$, there exists an edge $vw\in E(G)$ with $v\in V_x$ and $w\not\in V_x$.  Since $v\in V_x$, $v\in Y_{x'}$ for some $T$-descendant $x'$ of $x$ (possibly $x=x'$). Since $\mathcal{Y}$ is a partition, $w\in Y_{a}$ for some $a\not\in V(T_x)$.  Since $vw\in E(G)$, we have $x'a\in E(H)$.  By \tref{ancestor-edge}, one of $a$ or $x'$ is a $T$-ancestor of the other. Since $w\in Y_a\subseteq V_a$ and $w\not\in V_x\supseteq V_{x'}$, \yref{containment-i} rules out the possibility that $x'$ is a $T$-ancestor of $a$. Therefore, $a$ is a $T$-ancestor of $x$ which is a $T$-ancestor of $x'$.  Let $z_0,\ldots,z_r$ be the path in $T$ from $z_0:=x'$ to $z_r:=a$.  Since $x'a\in E(H)$, at least one of $a$ or $x'$ is in $B_{z_i}$, for each $i\in\{0,\ldots,r\}$ and there is at least one $i$ such that $B_{z_i}$ contains both $x'$ and $a$.  However, by \tref{subtree-root} $x'$ is not contained in $B_{z_i}$ for any $i\in\{1,\ldots,r\}$.  Therefore $a\in B_{z_i}$ for each $i\in\{0,\ldots,r\}$.  In particular, $a$ is contained in $B_x$.
  Property~\yref{ancestor-edge} now follows from the fact that $|B_x|\le t+1$ and $B_x$ contains $x$.
\end{proof}

% \referee{2}{Page 7.  In the very last sentence, I think $B_{x_i}$ should be $B_{z_i}$
% (twice).  Moreover, the way the proof is written, it seems to suggest
% that we only get that $a \in B_{z_i}$ for each $i \in \{1, \dots, r\}$.
% The conclusion that $a \in B_{z_i}$ for each $i \in \{0, \dots, r\}$ is
% true though.  In a normalized tree-decomposition, if $xy$ is an edge and
% $x$ is a $T$-ancestor of $y$, then $x$ must be in $B_y$ (since the subtrees $T_x$
% and $T_y$ intersect).}

%\note{PM}{I addressed the preceding referee's.}

@@

\note{PM}{Same comment: Yes, these claims are definition checking.  We'll see if the experimental notation change helps.}

@@

\note{PM}{I also found the steps in this proof tedious. Unfortunately, when I tried to give a simple explanation for it in Banff, both Piotr and Torsten remained unconvinced until we opened the paper and walked through the steps of the proof together.  I suspect the referee who can see through these proofs if one of the authors of \cite{PS21}.}

@@

% \referee{2}{Page 18.  Say $H'$ is connected since $X$ is connected.}
% \note{PM}{This is in the wrong place. It refers to the proof about centered colouring, which is now just a reference.}

@@

\note{DW}{Say something about how \cref{1_planar_is_4_framed} is better than similar results in the literature.}
\note{PM}{I don't know what to say without explaining that $4$-framed (simple) graphs only contain $3$-connected $1$-planar graphs.  Vida's example shows that there is a $1$-plane embedding of a graph that can't be augmented into a $4$-framed graph, but I haven't checked if it can't be drawn some better way. And I also don't know if the Bekos claim that $8$-framed (smple) graphs contain all $1$-planar graphs is proven anywhere and, if so, is it tight?}

@@

\begin{thm}
\label{gkPlanarProduct}
Every $(g,k)$-planar graph is contained in $H\boxtimes P \boxtimes K_\ell$ for some graph $H$ with $\tw(H) \leq \binom{k+4}{3}-1$, where $\ell:=\max\{2g,3\}\cdot(6k^2+16k+10)$.
% \note{DW}{I have changed $\tw(H) \leq \binom{k+5}{4}-1$ to $\tw(H) \leq \binom{k+4}{3}-1$, okay?}
%\note{PM}{Ok}
\end{thm}

\note{DW}{I am pretty sure one can get treewidth 3 for $(g,1)$-planar graphs, combining \cref{framed} with the method in \citep{DHHW}. I think the best place for this result is reference \citep{DHHW} since the present paper pre-dates \citep{DHHW} (and this addition will increase the significance of \citep{DHHW}). Okay?}
\note{PM}{Okay to this and to all similar questions that come later (which I've now commented out.)}

@@

\note{DW}{I don't want to do a complete restructuring. In our response, let's just say why we structure it the way we do.}
\note{PM}{Sure, though I think they proposing a chop rather than a restructuring.}

@@

\cref{qn,gkPlanarProduct} imply that $(g,k)$-planar graphs have queue-number at most $g 2^{O(k^3)}$. 
% \note{DW}{I have changed $g 2^{O(k^4)}$ to $g 2^{O(k^3)}$, okay?}
% \note{OM}{Ok}
Note that \citet{DJMMUW20} previously proved the bound of
$O(g^{k+2})$ using \cref{GenusProduct} and an ad-hoc method. Our result provides a better bound when $g>2^{k^2}$ 
% \note{DW}{I have changed $g>2^{k^3}$ to $g>2^{k^2}$, okay?}. 
% \note{PM}{Ok}
In the case of 1-planar graphs we can improve further. \citet{ABGKP20} proved that every planar graph with treewidth at most $3$ has queue-number at most $5$. Thus the graph $H$ in \cref{1-planar} has queue-number at most $5$. \cref{qn,1-planar} then imply:

@@

\begin{cor}\quad
\begin{compactitem}
\item	For every $1$-planar graph $G$, $\pi(G)\le 7\times 4^4=1792$.
%\item For every $k$-planar graph $G$,
%	$\pi(G)\le (18k^2+48k+30) 4^{\binom{k+4}{3}}$.
\item	For every $(g,k)$-planar graph $G$,
	$  \pi(G)\le \max\{2g,3\}\cdot(6k^2+16k+10) 4^{\binom{k+4}{3}}.$
% \note{DW}{I have changed $4^{\binom{k+5}{4}}$ to $4^{\binom{k+4}{3}}$, okay?}
% \note{PM}{Ok}
\end{compactitem}
\end{cor}

@@

\begin{cor}\quad
\begin{compactitem}
\item For every $1$-planar graph $G$,\; $ \chi_p(G)\le 5 (p+3)(p+2)(p+1)^2$.
%\item For every $k$-planar graph $G$,\; $\displaystyle \chi_p(G)\le (18k^2+48k+30)(p+1) \binom{p+ \binom{k+4}{3}-1}{ \binom{k+4}{3}-1}$.
\item For every $(g,k)$-planar graph $G$,\;
$\displaystyle \chi_p(G) \le \max\{2g,3\}\cdot(6k^2+16k+10) (p+1) \binom{p+\binom{k+4}{3}-1}{\binom{k+4}{3}-1}$.
% \note{DW}{I have changed $\binom{k+5}{4}$ to $\binom{k+4}{3}$, okay?}
% \note{PM}{Ok}
 \end{compactitem}
\end{cor}

@@

\note{DW}{This subsection needs a clean-up; define $(g,d)$-framed graph. Is the following lemma known (at least in the $g=0$ case)?}
\note{PM}{I have no idea if this is known.  If so, it's probably not stated in terms of frames.}
\begin{lem}
\label{MewMapGraphLemma}
for all integers $g\geq 0$ and $d\geq 3$, every $(g,d)$-map graph is a spanning subgraph of  $(g,d)$-framed graph.
\end{lem}

@@

% \note{DW}{\cref{MapPartition} has been updated.}

% \note{DW}{I am pretty sure one can get treewidth 3 here using \cref{framed} and the method in \citep{DHHW}. I think the best place for this result is reference \citep{DHHW} since the present paper pre-dates \citep{DHHW} (and this addition will increase the significance of \citep{DHHW}). Okay?}

@@

% \note{DW}{\cref{StringPartition} has been updated.}

@@

% \note{DW}{\cref{PowerGenus} has been updated}
